A new method to estimate dark matter halo concentrations

### 2015-04-20 paper/article.tex
edits in the introduction

\section{Introduction}
\label{sec:introduction}
In the concordance cosmology paradigm the the matter content of the
Universe is dominated by dark matter which behaves as a collisionless
fluid under the influence of gravity. In the last three decades
numerical experiments have made possible the simulation of dark matter
dominated universes, yielding valuable insights on the large scale
structure formation process as depicted in observations
\citep{Springel2005,2011ApJ...740..102K}.

One of the most striking results of these simulations is that dark
matter clumps on galactic scales follow a universal density profile
which in a first approximation is spherical symmetric and only
dependent on the radial coordinate. These profiles seem to be to a
very good approximation universal \citep{NFW,Taylor2001}; independent
of the cosmological parameters and self-similar for different spatial
scales after an adequate re-scaling is applied.

One the most common parameterization of this density is known as the
Navarro-Frenk-White (NFW) profile \citep{NFW}. This profile is a
double power law in radius, where the transition between the two
happens at the so-called scale radius $r_s$. The ratio between the
scale radius and the virial radius of the halo $R_v$ is known as the
concentration $c=R_v/r_s$, a quantity that has been found to be
correlated with the halo mass.

In spite of the fact that high resolution simulations of
Milky Way sized dark matter halos \citep{Navarro2010} show that the
universality property is not  perfect and that a better fitting
parameterization to the radial density profile is provided by the
Einasto profile \citep{Einasto1965}, the NFW and its concentration
have become a standard and useful way to describe the structure of
dark matter halos.

The relationship between halo mass and concentration is in principle
accessible to observations and provides a potential test of LCDM on
galactic scales. 
With this promise a great deal of effort has been
invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} but also
finding the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

From the computational point of view there are two main methods to
estimate the concentration parameter of a dark matter halo in a N-body
simulation. The first method takes the particles composing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of radius. A second method uses an analytic property of the
NFW that related the maximum of the ratio of the circular velocity to
the virial velocity. In this method at each particle radius this ratio
is estimated to find the maximum value for this ratio; this value is
used to find the roots of an equation which represent the
concentration parameter.

The first method although is straightforward to apply has two main
disadvantages. First of all it requires a large number of particles in
order to have a proper density estimate in each bin. For a very low
number of particles is hard to define a large number of bins to
proceed with the fit. The second problem is that, as in any process
involving data binning, there is not an way to estimate the optimal
bin size, which in turn can affect the results of the fit.

The second method solves the two problems mentioned above. It works
with low particles numbers and does not involve data binning. However,
it effectively takes into account a single data point and discards the
behaviour of the ratio $V_{cirv}/V_{vir}$ below and above its
maxima.

>>>>

\section{Introduction}
\label{sec:introduction}
In the concordance cosmology paradigm the matter content of the
Universe is dominated by dark matter, a collisionless fluid shaped by
gravitational interactions.
In the last three decades numerical experiments have made possible the
simulation of dark matter dominated universes, yielding valuable
insights on the large scale structure formation process.
This description shows a remarkable success in comparisons against
observations of galaxy distributions on large scales
\citep{Springel2005,2011ApJ...740..102K}.

Regarding the small scale dark matter distribution, one of the most
striking results of these simulations is that dark matter overdensities
follow a universal density profile.
In a first approximation this profile is spherically symmetric and its
density only dependens on the radial coordinate. 
The universality of this profiles seems to be independent of the
cosmological parameters and self-similar for different spatial scales
after an adequate re-scaling is applied.  
\citep{NFW,Taylor2001}

One the most popular parameterization for a halo radial density
distribution is known as the Navarro-Frenk-White (NFW) profile
\citep{NFW}. 
This profile is a  double power law in radius, where the transition
between the two happens at the so-called scale radius $r_s$. 
The ratio between the scale radius and the virial radius of the halo
$R_v$ is known as the concentration $c=R_v/r_s$. 
Simulations also show that the concentration is a function of halo
mass.

High resolution simulations of Milky Way sized dark matter halos
\citep{Navarro2010} show that the universality property is not
perfect and that a better fitting parameterization to the radial
density profile is provided by the Einasto profile
\citep{Einasto1965}.
However, the NFW density profile and its concentration have become a
standard metric to describe the structure of dark matter
halos.

Observationally, the relationship between halo mass and concentration
could provides a potential test of LCDM on galactic scales. 
With this promise a great deal of effort has been
invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} but also
finding the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

From the computational point of view there are two main methods to
estimate the concentration parameter of a dark matter halo in a N-body
simulation. 
The first method takes the particles composing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of the radius. 
A second method uses an analytic property of the
NFW that relates the maximum of the ratio of the circular velocity to
the virial velocity. 
In this method the user estimates this ratio as a function of radius
to determine its maxium value.
Then, the user can take this value to find the roots of an equation
which represent the concentration parameter.

The first method is straightforward to apply, but it has two 
disadvantages. 
First, it requires a large number of particles in
order to have a proper density estimate in each bin. 
This makes the method robust only when halos with at least  $10^3$
particles are fitted.
The second problem is that there is not a way to estimate the optimal
bin size.
Different choices affect the results of the fit.

The second method solves the two problems mentioned above. 
It works with low particles numbers and does not involve data
binning. 
However, it effectively takes into account only a single data point
and discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below and
above its maxima. 
Additionally, small fluctuations on the value of this maximum can
yield large perturbations on the estimated concentration parameter.

%%%
